---
title: "Project_1_Normalization"
author: "Ashley Morrison"
output: html_document
date: "2023-02-07"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
setwd("/projectnb/bf528/users/group_7/project_1/")

#if (!require("BiocManager", quietly = TRUE))
#  install.packages("BiocManager")
#BiocManager::install(version = "3.16")

#BiocManager::install(c("affy","affyPLM","sva","AnnotationDbi","hgu133plus2.db"))

library(affy)
library(affyPLM)
library(sva)
library(AnnotationDbi)
library(hgu133plus2.db)
install.packages("factoextra")
install.packages("vctrs")
install.packages("ggplot2")
library(vctrs)
library(ggplot2)
library(factoextra)

```

## Normalization

Read in data (3.3)
```{r}
rawdata <- ReadAffy(celfile.path = "./samples") #read in CEL files
exps <- rma(rawdata) #background correcting, normalization, and generation expression for CEL files

```

Compute Relative Log Expression (RLE) and Normalized Unscaled Standard Error (NUSE) scores of the microarray samples (3.4)
info on package: https://bioconductor.org/packages/release/bioc/manuals/affyPLM/man/affyPLM.pdf

QUESTION! Which plot to use to visualize
```{r}
data_PLM <- fitPLM(object = rawdata,normalize=TRUE,background = TRUE)
#image(data_PLM)
NUSE(data_PLM,xaxt = "n",ylim=c(.9,2.2))
title(main="Normalized Unscaled Standard Error (NUSE)",xlab = "Sample", ylab = "NUSE")

NUSE(data_PLM,xaxt = "n",ylim=c(.9,1.1))
title(main="Normalized Unscaled Standard Error (NUSE), Zoomed in",xlab = "Sample", ylab = "NUSE")


#add in plot info
RLE(data_PLM,xaxt = "n",ylim=c(-10,10)) #add in plot info
title(main="Relative Log Expression (RLE)",xlab = "Sample", ylab = "RLE")

RLE(data_PLM,xaxt = "n") #add in plot info
title(main="Relative Log Expression (RLE), Zoomed in",xlab = "Sample", ylab = "RLE")

```


part 3.5

normalizationcombatbatch --> adjustment variables
normalizationcombatmod --> features of interest

QUESTIONS --> what to read into combat (raw vs normalized), what to treat each variable as
```{r}
metadata <- read.csv("./code/proj_metadata.csv")
head(metadata)

#read just expression data (USING normalized data)
edata_rma <- exprs(exps)

#checking that the order of samples is the same between expression and metadata
substr(colnames(edata_rma), 1, 9) == metadata$geo_accession

#defining model:
modcombat <- model.matrix(~as.factor(normalizationcombatmod),data = metadata)
batch <- metadata$normalizationcombatbatch

#Run ComBat
combat_edata <- ComBat(dat=edata_rma, batch=batch, mod=modcombat)

#write to csv for Analyst role:
#write.csv(combat_edata, file="proj1_intensity_data.csv")
```
